---
title: "Project 1 Redwood Data Code Appendix"
author: "Yicheng Shen (Student ID: 2806571) & Yunhong Bao (Student ID: 2427527)"
date: "October 13, 2022"
header-includes: 
      - \usepackage{amsmath}
output: 
    pdf_document    
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, warning=F)
library(mosaic)
library(GGally)
library(caret)
library(ggfortify)
library(ggbiplot)
library(gridExtra)
library(kableExtra)
library(pracma)
library(Hmisc)
library(lubridate)
library(mclust)
ggplot2::theme_set(ggplot2::theme_bw())
knitr::opts_chunk$set(out.width = "100%", fig.align = 'center')
```


# Data Cleaning

```{r read data, cache=T}
sonoma_data_all <- read.csv("data/sonoma-data-all.csv")
sonoma_data_log <- read.csv("data/sonoma-data-log.csv")
sonoma_data_net <- read.csv("data/sonoma-data-net.csv")
mote_location_data <- read.table("data/mote-location-data.txt", header = TRUE)
```


```{r summary data}
summary(sonoma_data_all)
summary(sonoma_data_net)
summary(sonoma_data_log)
```

```{r hist of each variable in raw data, message = F}
hist.data.frame(sonoma_data_all%>%
                    dplyr::select(c(voltage, humidity, humid_temp, humid_adj, hamatop, hamabot)))
```


```{r check duplicate data, eval = F, warning = F}
sonoma_data_all %>% 
  select(-c(voltage,result_time)) %>% 
  distinct() %>% 
  group_by(epoch, nodeid) %>% 
  dplyr::summarize(count = n()) %>%
  filter(count!=1)

sonoma_data_all %>% 
  filter(nodeid == 136, epoch == 17)

# sonoma_data_all %>% 
#   select(c(epoch,nodeid)) %>% 
#   distinct()
```

`2004-05-07 18:24:58` corresponds to `epoch` 2812

```{r retrieve time in sonoma_data_log}
epoch_2812_time <- c("2004-05-07 18:24:58.666424")

epoch_2812_time <- as.POSIXct(epoch_2812_time, "US/Pacific")

sonoma_data_log_with_time <- sonoma_data_log %>%
                                  mutate(real_time = epoch_2812_time - (2812-epoch)*60*5)
```

`voltage` conversion formula by ADC: $$\frac{1023}{X} = \frac{ADC}{0.580567}$$

```{r convert voltage in sonoma_data_net}
# table(sonoma_data_net$voltage)
# table(sonoma_data_log$voltage)

sonoma_data_net_real_vol <- sonoma_data_net %>% 
                              mutate(voltage = 1023*0.580567/voltage,      # ADC conversion
                                      real_time = as.POSIXct(result_time, "US/Pacific")) # cover time

# table(sonoma_data_net_real_vol$voltage) # double check
```


```{r conmbine data set, and adjust time convert hama}
all_data <- rbind(sonoma_data_log_with_time, sonoma_data_net_real_vol)

all_data <- all_data %>%
                distinct(epoch, nodeid, .keep_all = T)   # keep the first one of duplicate readings

all_data <- left_join(all_data, mote_location_data, by = c("nodeid" = "ID")) %>%
                    mutate(incident_par = hamatop / 54, reflected_par = hamabot / 54) # convert hama to PAR

all_data$real_time <- all_data$real_time - 7*60*60   # weird pitfall!!! result_time in net is wrong by 7 hours
```


```{r check outliers, out.width="75%", fig.align='center', warning = F}
a <- all_data %>% 
  ggplot() + geom_line(aes(real_time, humid_temp, color = factor(nodeid))) + 
  ylim(0, 100) + theme_bw() + theme(legend.position="none")

b <- all_data %>% 
  ggplot() + geom_line(aes(real_time, humidity, color = factor(nodeid))) + 
    ylim(-10, 130) +
  theme_bw() + theme(legend.position="none")

c <- all_data %>% 
  ggplot() + geom_line(aes(real_time, voltage, color = factor(nodeid))) + 
      ylim(0.5, 3.1) +
  theme_bw() + theme(legend.position="none")

grid.arrange(a,b,c,nrow=3)


all_data_clean <- all_data %>% 
                            na.omit() %>%                # Remove missing data
                            filter(voltage >= 2.4) %>%   # Remove battery failure
                            filter(humidity <= 100)      # Remove unreasonable humidity

a <- all_data_clean %>% 
  ggplot() + geom_line(aes(real_time, humid_temp, color = factor(nodeid))) + theme_bw() + theme(legend.position="none")

b <- all_data_clean  %>%
  ggplot() + geom_line(aes(real_time, humidity, color = factor(nodeid))) + theme_bw() + theme(legend.position="none")

c <- all_data_clean %>% 
  ggplot() + geom_line(aes(real_time, voltage, color = factor(nodeid))) + theme_bw() + theme(legend.position="none")

grid.arrange(a,b,c,nrow=3)
```

```{r visualization}
hist.data.frame(all_data_clean%>%
                    dplyr::select(c(humidity, humid_temp, humid_adj, voltage, incident_par, reflected_par)))
```

```{r remove additional outliers}
hist(all_data_clean$incident_par, breaks = 30)

summary(all_data_clean$incident_par)
summary(all_data_clean$reflected_par)

# all_data_clean %>% filter(incident_par > 2150)  # Problem with node 40 

a <- all_data_clean %>% filter(nodeid == 40) %>% 
  ggplot() + geom_line(aes(real_time, incident_par, color = factor(nodeid))) + theme_bw() + theme(legend.position="none")

b <- all_data_clean  %>% filter(nodeid == 40) %>% 
  ggplot() + geom_line(aes(real_time, reflected_par, color = factor(nodeid))) + theme_bw() + theme(legend.position="none")

c <- all_data_clean %>% filter(nodeid == 40) %>% 
  ggplot() + geom_line(aes(real_time, voltage, color = factor(nodeid))) + theme_bw() + theme(legend.position="none")

grid.arrange(a,b,c,nrow=3) # node 40 must be broken

all_data_clean <- all_data_clean %>% filter(nodeid != 40)

summary(all_data_clean) ## all good now
```


```{r save data for write-up plots, eval = F}
save(all_data, file = "data/all_data.RData")
save(all_data_clean, file = "data/all_data_clean.RData")
```



## Data Exploration 

Variables of interest: `humid_temp`, `humid_adj`, `incident_par`, `reflected_par`
Variables with node: `nodeid`, `epoch`, `voltage`
Variable with node location: `Height`, `Direc`, `Tree`


```{r all names}
names(all_data_clean)

all_data_clean <- all_data_clean %>% 
  mutate(`height level` = ifelse(Height > 50, "High", "Medium")) %>% 
  mutate(`height level` = ifelse(Height < 30, "Low", `height level`)) 
```

Choose consistent day(s)

```{r find a good time period}
all_data_clean %>% 
  mutate(Date = as.Date(all_data_clean$real_time, "US/Pacific")) %>% 
  group_by(Date) %>% 
  dplyr::summarize(`number of data point` = n()) %>% 
  ggplot(aes(Date,`number of data point`)) + geom_point() + geom_line()

three_good_day <- all_data_clean %>%
              mutate(Date = as.character(as.Date(all_data_clean$real_time, "US/Pacific") ))%>% 
              filter(Date %in% c("2004-04-30", "2004-05-01", "2004-05-02"))

one_day_data <- all_data_clean %>%
              mutate(Date = as.character(as.Date(all_data_clean$real_time, "US/Pacific") ))%>% 
              filter(Date %in% c("2004-05-15"))
```

```{r find a good node}
all_data_clean %>% 
  group_by(nodeid, `height level`) %>% 
  dplyr::summarize(`number of data point` = n()) %>% 
  ggplot() + geom_point(aes(nodeid,`number of data point`, color = `height level`), size = 2) +
   geom_text(aes(nodeid,`number of data point`, label=nodeid), size = 3) 
# good record from 42, 197, 14
```

1. 
Strong positive correlation: `humid_temp` & `voltage`
Strong negative correlation: `humid_temp` & `humid_adj`

```{r pairwise scatterplot, message=F}
three_good_day %>% 
  select(humid_temp, humid_adj, incident_par, reflected_par, voltage, Height) %>% 
  ggpairs() 

one_day_data %>% 
  select(humid_temp, humid_adj, incident_par, reflected_par, voltage, Height) %>% 
  ggpairs()
```

2. `incident_par` positively correlated with `humid_temp` and `reflected_par`

3. Time Series

```{r time seies plot}
temp_height <- ggplot(one_day_data) + geom_line(aes(real_time, humid_temp, col = `height level`))
humid_height <- ggplot(one_day_data) + geom_line(aes(real_time, humid_adj, col = `height level`))
inc_height <- ggplot(one_day_data) + geom_line(aes(real_time, incident_par, col = `height level`))
ref_height <- ggplot(one_day_data) + geom_line(aes(real_time, reflected_par, col = `height level`))
grid.arrange(temp_height, humid_height, inc_height,ref_height, nrow =2)

temp_height <- ggplot(three_good_day) + geom_line(aes(real_time, humid_temp, col = `height level`))
humid_height <- ggplot(three_good_day) + geom_line(aes(real_time, humid_adj, col = `height level`))
inc_height <- ggplot(three_good_day) + geom_line(aes(real_time, incident_par, col = `height level`))
ref_height <- ggplot(three_good_day) + geom_line(aes(real_time, reflected_par, col = `height level`))
grid.arrange(temp_height, humid_height, inc_height,ref_height, nrow =2)

temp_height <- ggplot(all_data_clean) + geom_line(aes(real_time, humid_temp, col = `height level`))
humid_height <- ggplot(all_data_clean) + geom_line(aes(real_time, humid_adj, col = `height level`))
inc_height <- ggplot(all_data_clean) + geom_line(aes(real_time, incident_par, col = `height level`))
ref_height <- ggplot(all_data_clean) + geom_line(aes(real_time, reflected_par, col = `height level`))
grid.arrange(temp_height, humid_height, inc_height,ref_height, nrow =2)
```


```{r plot by direction and tree}
ggplot(all_data_clean) + geom_line(aes(real_time, humid_temp)) + facet_wrap(~Direc)
ggplot(all_data_clean) + geom_line(aes(real_time, humid_temp)) + facet_wrap(~Tree)
```


```{r three node plot}
all_data_clean_3_nodes <- all_data_clean %>% 
  filter(nodeid %in% c(42, 197, 14))
  # filter(nodeid %in% c(42))

temp_height <- ggplot(all_data_clean_3_nodes) + geom_line(aes(real_time, humid_temp, col = `height level`))
humid_height <- ggplot(all_data_clean_3_nodes) + geom_line(aes(real_time, humid_adj, col = `height level`))
inc_height <- ggplot(all_data_clean_3_nodes) + geom_line(aes(real_time, incident_par, col = `height level`))
ref_height <- ggplot(all_data_clean_3_nodes) + geom_line(aes(real_time, reflected_par, col = `height level`))
grid.arrange(temp_height, humid_height, inc_height,ref_height, nrow =2)
```



4. PCA

```{r PCA eval = F}
# PCA_result <- prcomp(all_data_clean %>% select("humid_temp","humid_adj", "incident_par", "reflected_par", "Height") ,center = TRUE, scale. = TRUE)

PCA_result <- prcomp(all_data_clean[c(8:9, 17:18)],center = TRUE, scale. = TRUE)
summary(PCA_result)
PCA_result

autoplot(PCA_result, label = F, label.size = 3,
         loadings = TRUE, loadings.label = TRUE, 
         loadings.label.size  = 3) + theme_bw()


ggbiplot(pcobj = PCA_result, choices = c(1,2),
                  obs.scale = 1, var.scale = 1,  # Scaling of axis
                  labels.size = 4, varname.size = 5,
                  varname.abbrev = TRUE,  # Abbreviate variable names (TRUE)
                  var.axes = TRUE,      # Remove variable vectors (TRUE)
                  circle = TRUE,        # Add unit variance circle (TRUE)
                  ellipse = FALSE, groups = as.character(as.Date(all_data_clean$real_time, "PDT")))

table(all_data_clean$Height)
```

```{r Scree plot}
eigenvalues <- (PCA_result$sdev)^2
eigs_cum <- cumsum(eigenvalues)/sum(eigenvalues)
a <- ggplot() + geom_point(aes(x = 1:length(eigenvalues), y=eigenvalues)) +
  geom_line(aes(x = 1:length(eigenvalues), y=eigenvalues)) +
  labs(x = "first PCs", y = "eigenvalues") + theme_bw()
b <- ggplot() + geom_point(aes(x = 1:length(eigenvalues), y=eigs_cum)) +
geom_line(aes(x = 1:length(eigenvalues), y=eigs_cum)) + theme_bw() +
  labs(x = "first PCs", y = "fraction of total variance explained")
grid.arrange(a,b, nrow = 1)
```

```{r PCA one day plot}
one_day_data_3h <- one_day_data %>% 
  mutate(date = as.Date(real_time, "US/Pacific"), hour = format((one_day_data$real_time), format = "%H") ) %>% 
  filter(hour %in% c("00", "12"))

PCA_result_1_day <- prcomp(one_day_data_3h[c(8:9, 17:18)],center = TRUE, scale. = TRUE)
summary(PCA_result_1_day)
PCA_result_1_day

# one_day_data$daytime <- ifelse(one_day_data$real_time >= "2004-05-15 07:00:00 PDT", "Day","Dark")
# one_day_data$daytime <- ifelse(one_day_data$real_time >= "2004-05-15 18:00:00 PDT", "Dark",one_day_data$daytime)

autoplot(PCA_result_1_day, label = F, label.size = 3,
         loadings = TRUE, loadings.label = TRUE, 
         loadings.label.size  = 3) + theme_bw()

ggbiplot(pcobj = PCA_result_1_day, choices = c(1,2),
                  obs.scale = 1, var.scale = 1,  # Scaling of axis
                  labels.size = 4, varname.size = 5,
                  varname.abbrev = TRUE,  # Abbreviate variable names (TRUE)
                  var.axes = TRUE,      # Remove variable vectors (TRUE)
                  circle = TRUE,        # Add unit variance circle (TRUE)
                  ellipse = FALSE, groups = one_day_data_3h$hour)

```

```{r one day clustering}
PCA_result_1_day_dat <- as_tibble(PCA_result_1_day$x[,1:2])


cl_km <- kmeans(PCA_result_1_day_dat, 2, nstart = 25)
z_km <- cl_km$cluster
kmeans_center <- cl_km$centers
k_means_plot <- ggplot() + geom_point(aes(PCA_result_1_day$x[, 1], PCA_result_1_day$x[, 2], col=factor(z_km)), size = 0.6) + labs(x = "PC1", y = "PC2", title = "K-means Algorithm", color = "K-mean Cluster")

model_GMM <- Mclust(PCA_result_1_day_dat, G = 2)
z_GMM = apply(model_GMM$z, 1, which.max)
GMM_means <- model_GMM$parameters$mean
EM_plot <- ggplot() + geom_point(aes(PCA_result_1_day$x[, 1], PCA_result_1_day$x[, 2], col=factor(z_GMM)), size = 0.6) + labs(x = "PC1", y = "PC2", title = "GMM Using EM", color = "GMM Cluster")

ggplot(PCA_result_1_day_dat) + 
  geom_point(aes(PC1, PC2, color = one_day_data_3h$hour), size = 0.6) 

k_means_plot
EM_plot
```


```{r PCA multiple day}
three_good_day_3h <- all_data_clean %>%
          mutate(date = as.character(as.Date(real_time, "US/Pacific")), hour = format((all_data_clean$real_time), format = "%H") ) %>% 
              filter(date %in% c("2004-05-02", "2004-05-17")) 

# all_data_clean %>%
#       mutate(date = as.character(as.Date(real_time, "US/Pacific")), hour = format((all_data_clean$real_time), format = "%H") ) %>% 
#               filter(date %in% c("2004-05-02")) %>%
#               ggplot() + geom_line(aes(real_time, humid_adj, col = `height level`))


PCA_result_3_day <- prcomp(three_good_day_3h[c(8:9, 17:18)],center = TRUE, scale. = TRUE)
summary(PCA_result_3_day)
PCA_result_3_day

autoplot(PCA_result_3_day, label = TRUE, label.size = 3,
         loadings = TRUE, loadings.label = TRUE, 
         loadings.label.size  = 3) + theme_bw()

ggbiplot(pcobj = PCA_result_3_day, choices = c(1,2),
                  obs.scale = 1, var.scale = 1,  # Scaling of axis
                  labels.size = 4, varname.size = 5,
                  varname.abbrev = TRUE,  # Abbreviate variable names (TRUE)
                  var.axes = TRUE,      # Remove variable vectors (TRUE)
                  circle = TRUE,        # Add unit variance circle (TRUE)
                  ellipse = FALSE, groups = three_good_day_3h$date)
```


```{r 3 day clustering}
PCA_result_3_day_dat <- as_tibble(PCA_result_3_day$x[,1:2])


cl_km <- kmeans(PCA_result_3_day_dat, 2, nstart = 25)
z_km <- cl_km$cluster
kmeans_center <- cl_km$centers
k_means_plot <- ggplot() + geom_point(aes(PCA_result_3_day$x[, 1], PCA_result_3_day$x[, 2], col=factor(z_km)), size = 0.6) + labs(x = "PC1", y = "PC2", title = "K-means Algorithm", color = "K-mean Cluster")

model_GMM <- Mclust(PCA_result_3_day_dat, G=2)
z_GMM = apply(model_GMM$z, 1, which.max)
GMM_means <- model_GMM$parameters$mean
EM_plot <- ggplot() + geom_point(aes(PCA_result_3_day$x[, 1], PCA_result_3_day$x[, 2], col=factor(z_GMM)), size = 0.6) + labs(x = "PC1", y = "PC2", title = "GMM Using EM", color = "GMM Cluster")

ggplot(PCA_result_3_day_dat) + 
  geom_point(aes(PC1, PC2, color = three_good_day_3h$date), size = 0.6) 
k_means_plot
EM_plot

# ggplot(PCA_result_3_day_dat) + 
#   geom_point(aes(PC1, PC2, color = three_good_day_3h$date), size = 0.6) 
```

```{r two node pca}
two_node_data <- all_data_clean %>%
          mutate(date = as.character(as.Date(real_time, "US/Pacific")), hour = format((all_data_clean$real_time), format = "%H") ) %>% 
              filter(hour %in% c("12")) %>% 
              filter(nodeid %in% c(14, 42)) 

PCA_two_node <- prcomp(two_node_data[c(8:9, 17:18)],center = TRUE, scale. = TRUE)
summary(PCA_two_node)
PCA_two_node


ggbiplot(pcobj = PCA_two_node, choices = c(1,2),
                  obs.scale = 1, var.scale = 1,  # Scaling of axis
                  labels.size = 4, varname.size = 5,
                  varname.abbrev = TRUE,  # Abbreviate variable names (TRUE)
                  var.axes = TRUE,      # Remove variable vectors (TRUE)
                  circle = TRUE,        # Add unit variance circle (TRUE)
                  ellipse = FALSE, groups = factor(two_node_data$nodeid))


```



```{r 2 node clustering}
PCA_two_node_dat <- as_tibble(PCA_two_node$x[,1:2])

cl_km <- kmeans(PCA_two_node_dat, 2, nstart = 25)
z_km <- cl_km$cluster
kmeans_center <- cl_km$centers
k_means_plot <- ggplot() + geom_point(aes(PCA_two_node$x[, 1], PCA_two_node$x[, 2], col=factor(z_km)), size = 0.6) + labs(x = "PC1", y = "PC2", title = "K-means Algorithm", color = "K-mean Cluster")

model_GMM <- Mclust(PCA_two_node_dat, G=2)
z_GMM = apply(model_GMM$z, 1, which.max)
GMM_means <- model_GMM$parameters$mean
EM_plot <- ggplot() + geom_point(aes(PCA_two_node$x[, 1], PCA_two_node$x[, 2], col=factor(z_GMM)), size = 0.6) + labs(x = "PC1", y = "PC2", title = "GMM Using EM", color = "GMM Cluster")

ggplot(PCA_two_node_dat) + 
  geom_point(aes(PC1, PC2, color = factor(two_node_data$nodeid)), size = 0.6) 
k_means_plot
EM_plot

table(two_node_data$Height)
```


### Graph critiques

```{r log of 3a}
hist.data.frame(all_data_clean%>%
                  mutate(log_temp =log(humid_temp), log_humid = log(humid_adj), 
                         log_incident_par = log(incident_par), log_reflected_par = log(reflected_par)) %>% 
                    dplyr::select(c(log_temp, log_humid, log_incident_par, log_reflected_par)))

# hist(log(all_data_clean$incident_par))
```
```{r adjust 3c}
all_data_clean %>% filter(as.Date(real_time, "US/Pacific") == "2004-05-01") %>% 
  group_by(factor(Height)) %>% dplyr::summarize(mean_temp = mean(humid_temp)) %>% 
  ggplot(aes(x = as.numeric(`factor(Height)`), y = mean_temp)) + geom_point() + geom_smooth(se = F) + 
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.5)) 


all_data_clean %>% filter(as.Date(real_time, "US/Pacific") == "2004-05-01") %>% 
  group_by(factor(Height)) %>% dplyr::summarize(mean_humid = mean(humid_adj)) %>% 
  ggplot(aes(x = as.numeric(`factor(Height)`), y = mean_humid)) + geom_point() + geom_smooth(se = F) + 
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.5)) 
  # geom_hline(yintercept = mean(all_data_clean$humid_temp)) + coord_flip()



all_data_clean %>% filter(as.Date(real_time, "US/Pacific") == "2004-05-01") %>% 
  ggplot(aes(x = real_time, y = incident_par)) + geom_point(alpha = 0.7) + geom_smooth(se = F)
```


```{r adjust figrue 7}

yield_compare <- rbind(sonoma_data_log_with_time, sonoma_data_net_real_vol) 

yield_compare %>% 
  mutate(source = ifelse(result_time == "2004-11-10 14:25:00", "log", "net")) %>% 
  mutate(date = as.Date(real_time, "US/Pacific")) %>% 
  group_by(date, source) %>% 
  dplyr::summarize(`number of data yield` = n()) %>% 
  ggplot(aes(date, `number of data yield`)) + geom_point() + geom_line() + labs(x = "") + facet_wrap(~source)
# + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.5)) 

yield_compare %>% 
  mutate(source = ifelse(result_time == "2004-11-10 14:25:00", "log", "net")) %>% 
  mutate(date = as.Date(real_time, "US/Pacific"))  %>% 
  ggplot(aes(date)) + geom_bar()+ facet_wrap(~source)
  
  
left_join(yield_compare, mote_location_data, by = c("nodeid" = "ID")) %>% 
  na.omit() %>% 
  mutate(source = ifelse(result_time == "2004-11-10 14:25:00", "log", "net")) %>% 
  mutate(date = as.Date(real_time, "US/Pacific")) %>% 
  group_by(Height, source) %>% 
  dplyr::summarize(`number of data yield` = n()) %>% 
  ggplot(aes(Height, `number of data yield`)) + geom_point() + geom_line()+ labs(x = "") + facet_wrap(~source)

left_join(yield_compare, mote_location_data, by = c("nodeid" = "ID")) %>% 
    na.omit() %>% 
  mutate(source = ifelse(result_time == "2004-11-10 14:25:00", "log", "net")) %>% 
  mutate(date = as.Date(real_time, "US/Pacific")) %>% 
  ggplot(aes(as.factor(Height))) + geom_bar()+ facet_wrap(~source) +
     theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.5))  + coord_flip() + 
  labs(x = "Height", y = "Data Yield")
```



## backup below; don't run

```{r mean plot}
a <- all_data_clean %>% 
  mutate(date = as.Date(real_time, "US/Pacific"), hour = format((all_data_clean$real_time), format = "%H") ) %>% 
  group_by(date, hour) %>%
  dplyr::summarize(avg.temp = mean(humid_temp), avg.humid = mean(humid_adj)) %>% 
  mutate(time = as.POSIXct(paste(date, hour), format = "%Y-%m-%d %H")) %>% 
  ggplot() + geom_line(aes(time, avg.temp)) 

b <- all_data_clean %>% 
  mutate(date = as.Date(real_time, "US/Pacific"), hour = format((all_data_clean$real_time), format = "%H") ) %>% 
  group_by(date, hour) %>%
  dplyr::summarize(avg.temp = mean(humid_temp), avg.humid = mean(humid_adj)) %>% 
  mutate(time = as.POSIXct(paste(date, hour), format = "%Y-%m-%d %H")) %>% 
  ggplot() + geom_line(aes(time, avg.humid)) 

grid.arrange(a,b,nrow=2)
```


```{r daily mean data}
daily_data <- all_data_clean %>% 
  mutate(date = as.Date(real_time, "US/Pacific"), hour = format((all_data_clean$real_time), format = "%H") ) %>% 
  group_by(date) %>%
  dplyr::summarize(avg.temp = mean(humid_temp), avg.humid = mean(humid_adj), max_inc = max(incident_par), max_ref = max(reflected_par)) 

PCA_result_daily <- prcomp(daily_data[c(2:5)], center = TRUE, scale. = TRUE)
summary(PCA_result_daily)
PCA_result_daily

autoplot(PCA_result_daily, label = F, label.size = 3,
         loadings = TRUE, loadings.label = TRUE, 
         loadings.label.size  = 3) + theme_bw()

ggbiplot(pcobj = PCA_result_daily, choices = c(1,2),
                  obs.scale = 1, var.scale = 1,  # Scaling of axis
                  labels.size = 1.5, varname.size = 5,
                  varname.abbrev = TRUE,  # Abbreviate variable names (TRUE)
                  var.axes = TRUE,      # Remove variable vectors (TRUE)
                  circle = TRUE,        # Add unit variance circle (TRUE)
                  ellipse = FALSE,
                labels = daily_data$date )


PCA_result_daily_dat <- as_tibble(PCA_result_daily$x[,1:2])

ggplot(PCA_result_daily_dat) + 
  geom_point(aes(PC1, PC2, color = daily_data$avg.humid), size = 1) 

cl_km <- kmeans(PCA_result_daily_dat, 2, nstart = 25)
z_km <- cl_km$cluster
kmeans_center <- cl_km$centers
k_means_plot <- ggplot() + geom_point(aes(PCA_result_daily$x[, 1], PCA_result_daily$x[, 2], col=factor(z_km)), size = 1.5) + labs(x = "PC1", y = "PC2", title = "K-means Algorithm", color = "K-mean Cluster")

model_GMM <- Mclust(PCA_result_daily_dat, G=2)
z_GMM = apply(model_GMM$z, 1, which.max)
GMM_means <- model_GMM$parameters$mean
EM_plot <- ggplot() + geom_point(aes(PCA_result_daily$x[, 1], PCA_result_daily$x[, 2], col=factor(z_GMM)), size = 1.5) + labs(x = "PC1", y = "PC2", title = "GMM Using EM", color = "GMM Cluster")

k_means_plot
EM_plot

```



```{r, eval = F}
daily_2h_data <- all_data_clean %>% 
  mutate(date = as.Date(real_time, "US/Pacific"), hour = format((all_data_clean$real_time), format = "%H") ) %>% 
  filter(hour %in% c( "00"))

ggplot(daily_2h_data) + geom_boxplot(aes(factor(date), humid_temp))

ggplot(daily_2h_data) + geom_point(aes(date, humid_temp, color = `height level`), size = 0.7)

PCA_result_hourly <- prcomp(daily_2h_data[c(8:9, 17:18)],center = TRUE, scale. = TRUE)
summary(PCA_result_hourly)
PCA_result_hourly

autoplot(PCA_result_hourly, label = TRUE, label.size = 3,
         loadings = TRUE, loadings.label = TRUE, 
         loadings.label.size  = 3) + theme_bw()

ggbiplot(pcobj = PCA_result_hourly, choices = c(1,2),
                  obs.scale = 1, var.scale = 1,  # Scaling of axis
                  labels.size = 4, varname.size = 5,
                  varname.abbrev = TRUE,  # Abbreviate variable names (TRUE)
                  var.axes = TRUE,      # Remove variable vectors (TRUE)
                  circle = TRUE,        # Add unit variance circle (TRUE)
                  ellipse = FALSE, groups = daily_2h_data$hour)

```


