---
title: "Project 1 Redwood Data Code Appendix"
author: "Yicheng Shen (Student ID: 2806571) & Yunhong Bao (Student ID: 2427527)"
date: "October 13, 2022"
header-includes: 
      - \usepackage{amsmath}
output: 
    pdf_document    
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, warning=F)
library(mosaic)
library(GGally)
library(caret)
library(ggfortify)
#library(ggbiplot)
library(gridExtra)
library(kableExtra)
library(pracma)
library(Hmisc)
library(lubridate)
ggplot2::theme_set(ggplot2::theme_bw())
knitr::opts_chunk$set(out.width = "100%", fig.align = 'center')
```


# Data Cleaning

```{r read data, cache=T}
sonoma_data_all <- read.csv("data/sonoma-data-all.csv")
sonoma_data_log <- read.csv("data/sonoma-data-log.csv")
sonoma_data_net <- read.csv("data/sonoma-data-net.csv")
mote_location_data <- read.table("data/mote-location-data.txt", header = TRUE)
```


```{r summary data}
summary(sonoma_data_all)
summary(sonoma_data_net)
summary(sonoma_data_log)
```

```{r hist of each variable in raw data, message = F}
hist.data.frame(sonoma_data_all%>%
                    dplyr::select(c(voltage, humidity, humid_temp, humid_adj, hamatop, hamabot)))
```


```{r check duplicate data, eval = F, warning = F}
sonoma_data_all %>% 
  select(-c(voltage,result_time)) %>% 
  distinct() %>% 
  group_by(epoch, nodeid) %>% 
  dplyr::summarize(count = n()) %>%
  filter(count!=1)

sonoma_data_all %>% 
  filter(nodeid == 136, epoch == 17)

# sonoma_data_all %>% 
#   select(c(epoch,nodeid)) %>% 
#   distinct()
```

`2004-05-07 18:24:58` corresponds to `epoch` 2812

```{r retrieve time in sonoma_data_log}
epoch_2812_time <- c("2004-05-07 18:24:58.666424")

epoch_2812_time <- as.POSIXct(epoch_2812_time)

sonoma_data_log_with_time <- sonoma_data_log %>%
                                  mutate(real_time = epoch_2812_time - (2812-epoch)*60*5)
```

`voltage` conversion formula by ADC: $$\frac{1023}{X} = \frac{ADC}{0.580567}$$

```{r convert voltage in sonoma_data_net}
# table(sonoma_data_net$voltage)
# table(sonoma_data_log$voltage)

sonoma_data_net_real_vol <- sonoma_data_net %>% 
                              mutate(voltage = 1023*0.580567/voltage,      # ADC conversion
                                      real_time = as.POSIXct(result_time)) # cover time

# table(sonoma_data_net_real_vol$voltage) # double check
```


```{r conmbine data set, and adjust time convert hama}
all_data <- rbind(sonoma_data_log_with_time, sonoma_data_net_real_vol)

all_data <- all_data %>%
                distinct(epoch, nodeid, .keep_all = T)   # keep the first one of duplicate readings

all_data <- left_join(all_data, mote_location_data, by = c("nodeid" = "ID")) %>%
                    mutate(incident_par = hamatop / 54, reflected_par = hamabot / 54) # convert hama to PAR

all_data$real_time <- all_data$real_time - 7*60*60   # weird pitfall!!! time in net is wrong by 7 hours
```


```{r check outliers, out.width="75%", fig.align='center', warning = F}
a <- all_data %>% 
  ggplot() + geom_line(aes(real_time, humid_temp, color = factor(nodeid))) + 
  ylim(0, 100) + theme_bw() + theme(legend.position="none")

b <- all_data %>% 
  ggplot() + geom_line(aes(real_time, humidity, color = factor(nodeid))) + 
    ylim(-10, 130) +
  theme_bw() + theme(legend.position="none")

c <- all_data %>% 
  ggplot() + geom_line(aes(real_time, voltage, color = factor(nodeid))) + 
      ylim(0.5, 3.1) +
  theme_bw() + theme(legend.position="none")

grid.arrange(a,b,c,nrow=3)


all_data_clean <- all_data %>% 
                            na.omit() %>%                # Remove missing data
                            filter(voltage >= 2.4) %>%   # Remove battery failure
                            filter(humidity <= 100)      # Remove unreasonable humidity

a <- all_data_clean %>% 
  ggplot() + geom_line(aes(real_time, humid_temp, color = factor(nodeid))) + theme_bw() + theme(legend.position="none")

b <- all_data_clean  %>%
  ggplot() + geom_line(aes(real_time, humidity, color = factor(nodeid))) + theme_bw() + theme(legend.position="none")

c <- all_data_clean %>% 
  ggplot() + geom_line(aes(real_time, voltage, color = factor(nodeid))) + theme_bw() + theme(legend.position="none")

grid.arrange(a,b,c,nrow=3)
```

```{r visualization}
hist.data.frame(all_data_clean%>%
                    dplyr::select(c(humidity, humid_temp, humid_adj, voltage, incident_par, reflected_par)))
```

```{r remove additional outliers}
hist(all_data_clean$incident_par, breaks = 30)

summary(all_data_clean$incident_par)
summary(all_data_clean$reflected_par)

# all_data_clean %>% filter(incident_par > 2150)  # Problem with node 40 

a <- all_data_clean %>% filter(nodeid == 40) %>% 
  ggplot() + geom_line(aes(real_time, humid_temp, color = factor(nodeid))) + theme_bw() + theme(legend.position="none")

b <- all_data_clean  %>% filter(nodeid == 40) %>% 
  ggplot() + geom_line(aes(real_time, humidity, color = factor(nodeid))) + theme_bw() + theme(legend.position="none")

c <- all_data_clean %>% filter(nodeid == 40) %>% 
  ggplot() + geom_line(aes(real_time, voltage, color = factor(nodeid))) + theme_bw() + theme(legend.position="none")

grid.arrange(a,b,c,nrow=3) # must be broken

all_data_clean <- all_data_clean %>% filter(nodeid != 40)

summary(all_data_clean) ## all good now
```


```{r save data for write-up plots, eval = F}
save(all_data, file = "data/all_data.RData")
save(all_data_clean, file = "data/all_data_clean.RData")
```



## Data Exploration 

Variables of interest: `humid_temp`, `humid_adj`, `incident_par`, `reflected_par`
Variables with node: `nodeid`, `epoch`, `voltage`
Variable with node location: `Height`, `Direc`, `Tree`


```{r all names}
names(all_data_clean)
```

Choose consistent day(s)

```{r find a good time period}
all_data_clean %>% 
  mutate(date = as.Date(all_data_clean$real_time)) %>% 
  group_by(date) %>% 
  dplyr::summarize(`number of data point` = n()) %>% 
  ggplot(aes(date,`number of data point`)) + geom_point() + geom_line()

three_good_day <- all_data_clean %>%
              mutate(date = as.character(as.Date(all_data_clean$real_time) ))%>% 
              filter(date %in% c("2004-04-30", "2004-05-01", "2004-05-02"))

one_day_data <- all_data_clean %>%
              mutate(date = as.character(as.Date(all_data_clean$real_time) ))%>% 
              filter(date %in% c("2004-05-15"))
```

1. 
Strong positive correlation: `humid_temp` & `voltage`
Strong negative correlation: `humid_temp` & `humid_adj`

```{r pairwise scatterplot, message=F}
three_good_day %>% 
  select(humid_temp, humid_adj, incident_par, reflected_par, voltage, Height) %>% 
  ggpairs() 

one_day_data %>% 
  filter(date == "2004-05-15") %>% 
  select(humid_temp, humid_adj, incident_par, reflected_par, voltage, Height) %>% 
  ggpairs()
```

2. `incident_par` positively correlated with `humid_temp` and `reflected_par`


```{r, eval = F, echo =F}
data_520 <- all_data_clean %>%
              mutate(date = as.character(as.Date(all_data_clean$real_time) )) %>% 
              filter(epoch == 1062)

table(data_520$real_time)

# one_day_data <- sonoma_data_net %>%
#               filter(result_time > "2004-05-22 23:00:59.075427", result_time < "2004-05-23 23:00:59.075427")
# one_day_data$real_time <- as.POSIXct(one_day_data$result_time)
# ggplot(one_day_data) + geom_line(aes(real_time, hamatop))
```


3. Time Series

```{r time seies plot}
all_data_clean_height <- all_data_clean %>% 
  mutate(`height level` = ifelse(Height > 50, "High", "Medium")) %>% 
  mutate(`height level` = ifelse(Height < 30, "Low", `height level`)) 

one_day_data_height <- one_day_data %>% 
  mutate(`height level` = ifelse(Height > 50, "High", "Medium")) %>% 
  mutate(`height level` = ifelse(Height < 30, "Low", `height level`)) 

three_good_day_height <- three_good_day %>% 
  mutate(`height level` = ifelse(Height > 50, "High", "Medium")) %>% 
  mutate(`height level` = ifelse(Height < 30, "Low", `height level`)) 

temp_height <- ggplot(one_day_data_height) + geom_line(aes(real_time, humid_temp, col = `height level`))
humid_height <- ggplot(one_day_data_height) + geom_line(aes(real_time, humid_adj, col = `height level`))
inc_height <- ggplot(one_day_data_height) + geom_line(aes(real_time, incident_par, col = `height level`))
ref_height <- ggplot(one_day_data_height) + geom_line(aes(real_time, reflected_par, col = `height level`))
grid.arrange(temp_height, humid_height, inc_height,ref_height, nrow =2)


temp_height <- ggplot(three_good_day_height) + geom_line(aes(real_time, humid_temp, col = `height level`))
humid_height <- ggplot(three_good_day_height) + geom_line(aes(real_time, humid_adj, col = `height level`))
inc_height <- ggplot(three_good_day_height) + geom_line(aes(real_time, incident_par, col = `height level`))
ref_height <- ggplot(three_good_day_height) + geom_line(aes(real_time, reflected_par, col = `height level`))
grid.arrange(temp_height, humid_height, inc_height,ref_height, nrow =2)

temp_height <- ggplot(all_data_clean_height) + geom_line(aes(real_time, humid_temp, col = `height level`))
humid_height <- ggplot(all_data_clean_height) + geom_line(aes(real_time, humid_adj, col = `height level`))
inc_height <- ggplot(all_data_clean_height) + geom_line(aes(real_time, incident_par, col = `height level`))
ref_height <- ggplot(all_data_clean_height) + geom_line(aes(real_time, reflected_par, col = `height level`))
grid.arrange(temp_height, humid_height, inc_height,ref_height, nrow =2)
```


```{r plot by direction and tree}
ggplot(all_data_clean_height) + geom_line(aes(real_time, humid_temp)) + facet_wrap(~Direc)
ggplot(all_data_clean_height) + geom_line(aes(real_time, humid_temp)) + facet_wrap(~Tree)
```


4. PCA

```{r PCA}
PCA_result <- prcomp(all_data_clean[c(8:9, 17:18)],center = TRUE, scale. = TRUE)
summary(PCA_result)
PCA_result
```

```{r Scree plot}
eigenvalues <- (PCA_result$sdev)^2
eigs_cum <- cumsum(eigenvalues)/sum(eigenvalues)
a <- ggplot() + geom_point(aes(x = 1:length(eigenvalues), y=eigenvalues)) +
  geom_line(aes(x = 1:length(eigenvalues), y=eigenvalues)) +
  labs(x = "first PCs", y = "eigenvalues") + theme_bw()
b <- ggplot() + geom_point(aes(x = 1:length(eigenvalues), y=eigs_cum)) +
geom_line(aes(x = 1:length(eigenvalues), y=eigs_cum)) + theme_bw() +
  labs(x = "first PCs", y = "fraction of total variance explained")
grid.arrange(a,b, nrow = 1)
```


```{r}
a <- all_data_clean %>% 
  mutate(date = as.Date(real_time), hour = format((all_data_clean$real_time), format = "%H") ) %>% 
  group_by(date, hour) %>%
  dplyr::summarize(avg.temp = mean(humid_temp), avg.humid = mean(humid_adj)) %>% 
  mutate(time = as.POSIXct(paste(date, hour), format = "%Y-%m-%d %H")) %>% 
  ggplot() + geom_line(aes(time, avg.temp)) 

b <- all_data_clean %>% 
  mutate(date = as.Date(real_time), hour = format((all_data_clean$real_time), format = "%H") ) %>% 
  group_by(date, hour) %>%
  dplyr::summarize(avg.temp = mean(humid_temp), avg.humid = mean(humid_adj)) %>% 
  mutate(time = as.POSIXct(paste(date, hour), format = "%Y-%m-%d %H")) %>% 
  ggplot() + geom_line(aes(time, avg.humid)) 

grid.arrange(a,b,nrow=2)
```







### data cleaning thoughts

```{r, eval = F, echo = F}
# do not touch 

flash_log_time <- c("2004-11-10 14:25:00")

sonoma_data_all %>%
  mutate(real_time = ifelse(result_time == flash_log_time, 
                            as.POSIXct(as.POSIXct("2004-05-07 18:24:58.666424") - (2812 - epoch)*60*5), 
                            as.POSIXct(result_time))) %>% 
  ggplot() + geom_line(aes(x = real_time, y = humidity, color = factor(nodeid)))

```





```{r}
hist.data.frame(sonoma_data_net%>%dplyr::select(-result_time,-epoch,-nodeid, -parent))
plot(sonoma_data_net$humidity, sonoma_data_net$voltage)
```


```{r}
hist(sonoma_data_net$voltage, breaks = 30)
summary(sonoma_data_net$voltage)
summary(sonoma_data_log$voltage)
```

```{r voltage problem}
table(sonoma_data_net$voltage)
sonoma_data_net$voltage

strange_node <- sonoma_data_net %>%
  filter(voltage == 1023 )

table(strange_node$nodeid)


table(sonoma_data_log$voltage)

strange_node_log <- sonoma_data_log %>%
  filter(voltage == 0.580567)

table(strange_node_log$nodeid)
```




```{r clean sonoma_data_log}

sonoma_data_log_clean <- sonoma_data_log %>% 
                            filter(voltage <= 3, voltage >= 2.4)

# hist(sonoma_data_log_clean$humid_temp)
# hist(sonoma_data_log_clean$humidity)

summary(sonoma_data_log_clean)

hist.data.frame(sonoma_data_log_clean%>%dplyr::select(-c(result_time, epoch, nodeid, parent, depth, real_time)))

# remove na
sonoma_data_log_clean2 <- sonoma_data_log_clean %>% na.omit() 

# check na
summary(sonoma_data_log_clean2)

# add location data
sonoma_data_log_clean3 <- left_join(sonoma_data_log_clean2, mote_location_data, by = c("nodeid" = "ID"))

hist.data.frame(sonoma_data_log_clean3%>%dplyr::select(c(humidity, humid_temp, hamatop, hamabot)))

log_clean_four <- sonoma_data_log_clean3%>%dplyr::select(c(humidity, humid_temp, hamatop, hamabot))
 
par(mfrow = c(2,2));
invisible(lapply(X=colnames(log_clean_four),FUN=function(s)boxplot(log_clean_four[,s], 
                main=paste("Histogram of", s),xlab=s, horizontal = T)))


sonoma_data_log_final <- sonoma_data_log_clean3
```


```{r clean net}
sonoma_data_net_clean <- sonoma_data_net %>% filter(voltage <= 500)

# adjust voltage
m_net <- mean(sonoma_data_net_clean$voltage)
m_log <- mean(sonoma_data_log_clean$voltage)
ratio <- m_net / m_log 

# filter voltage
sonoma_data_net_clean2 <- sonoma_data_net_clean %>%
                            mutate(voltage_new = voltage / ratio) %>% 
                            filter(voltage_new <= 3, voltage_new >= 2.5)

# check dist. of temp
plot(sonoma_data_net_clean2$humid_temp)

# plot hist for every var.
hist.data.frame(sonoma_data_net_clean2 %>% dplyr::select(-c(
  result_time, voltage, epoch, nodeid, parent, depth)))

# remove na
sonoma_data_net_clean3 <- sonoma_data_net_clean2 %>% na.omit() 

# summary(sonoma_data_net_clean3)

# add location data
sonoma_data_net_clean4 <- left_join(sonoma_data_net_clean3, mote_location_data, by = c("nodeid" = "ID"))

sonoma_data_net_final <- sonoma_data_net_clean4
```



```{r examine hama}
sonoma_data_log_hama <- sonoma_data_log_final %>%
                              mutate(incident_par = hamatop / 54, reflected_par = hamabot / 54)

summary(sonoma_data_log_hama)
hist(sonoma_data_log_hama$incident_par)
hist(sonoma_data_log_hama$reflected_par)

sonoma_data_net_hama <- sonoma_data_net_final %>%
                              mutate(incident_par = hamatop / 54, reflected_par = hamabot / 54)

summary(sonoma_data_net_hama)
hist(sonoma_data_net_hama$incident_par)
hist(sonoma_data_net_hama$reflected_par)
```










```{r Study of Temperature and Voltage in net, warning = F}
sonoma_data_net_norm_temp <- sonoma_data_net %>% filter(voltage<500)

plot( sonoma_data_net_norm_temp$voltage,
          sonoma_data_net_norm_temp$humid_temp)
```


```{r overall node failures}
sonoma_data_net_norm_temp$real_time <- as.POSIXct(sonoma_data_net_norm_temp$result_time)

a <- sonoma_data_net_norm_temp %>% 
  ggplot() + geom_line(aes(real_time, humid_temp, color = factor(nodeid))) + theme_bw()

b <- sonoma_data_net_norm_temp %>% 
  ggplot() + geom_line(aes(real_time, humidity, color = factor(nodeid))) + theme_bw()

c <- sonoma_data_net_norm_temp %>% 
  ggplot() + geom_line(aes(real_time, voltage, color = factor(nodeid))) + theme_bw()

grid.arrange(a,b,c,nrow=3)

# why voltage failure goes down??? 


sonoma_data_net %>% filter(voltage == 1023) %>% mutate(real_time = as.POSIXct(result_time)) %>%
  ggplot() + geom_line(aes(real_time, humid_temp, color = factor(nodeid))) + theme_bw()

sonoma_data_net %>% filter(voltage == 1023) %>% mutate(real_time = as.POSIXct(result_time)) %>%
  ggplot() + geom_line(aes(real_time, humidity, color = factor(nodeid))) + theme_bw()

```


```{r specific node failure}
a <- sonoma_data_net_norm_temp %>% 
  filter(nodeid == 74) %>% 
  ggplot() + geom_line(aes(real_time, humid_temp)) + theme_bw()

b <- sonoma_data_net_norm_temp %>% 
  filter(nodeid == 74) %>% 
  ggplot() + geom_line(aes(real_time, humidity)) + theme_bw()

c <- sonoma_data_net_norm_temp %>% 
  filter(nodeid == 74) %>% 
  ggplot() + geom_line(aes(real_time, voltage)) + theme_bw()

grid.arrange(a,b,c,nrow=3)



a <- sonoma_data_net_norm_temp %>%
    filter(nodeid == 197) %>% 
    ggplot() + geom_line(aes(real_time, humid_temp)) + theme_bw()
  
b <- sonoma_data_log_with_time  %>% 
    filter(nodeid == 197) %>% 
    ggplot() + geom_line(aes(real_time, humid_temp)) + theme_bw()


sonoma_data_log_with_time  %>% 
    filter(nodeid == 3) %>% select (epoch) %>% summary()

sonoma_data_net  %>% 
    filter(nodeid == 3) %>% select (epoch) %>% summary()
    
grid.arrange(a,b,nrow=2)
```




```{r}
# sonoma_data_log_with_time

min(sonoma_data_log_with_time$real_time)
max(sonoma_data_log_with_time$real_time)

a <- sonoma_data_log_voltage_clean %>% 
  ggplot() + geom_line(aes(real_time, humid_temp, color = factor(nodeid))) + theme_bw()

b <- sonoma_data_log_voltage_clean %>% 
  ggplot() + geom_line(aes(real_time, humidity, color = factor(nodeid))) + theme_bw()

c <- sonoma_data_log_voltage_clean %>% 
  ggplot() + geom_line(aes(real_time, voltage, color = factor(nodeid))) + theme_bw()

grid.arrange(a,b,c,nrow=3)
```

